Equation-of-state insensitive relations after GW170817

### 2019-03-17 paper/Submit/bin-love-after-gw170817.tex
arxive/prd version ready for submission
Added R-Love citations
adding Hectors references

\author{Zachary Carson}
\affiliation{%
 Department of Physics, University of Virginia, Charlottesville, Virginia 22904, USA
}%

\author{Katerina Chatziioannou}
%\affiliation{%
% Canadian Institute for Theoretical Astrophysics, 60 St. George Street, Toronto, Ontario, M5S 3H8, Canada
%}%
\affiliation{%
 Center for Computational Astrophysics, Flatiron Institute, 162 5th Ave, New York, NY 10010, USA
}%

@@

EoS-insensitive relations can in fact be used for more than just measuring the nuclear EoS. For years, the theoretical physics community considered the possibility of using measurements of NS properties, such as the mass, the radius and the moment of inertia, to constrain deviations from General Relativity in the strong-field regime. Certain modified theories of gravity, such as scalar tensor theories with spontaneous scalarization~\cite{Damour:1996ke}, Einstein-\AE ther and Ho\v rava gravity~\cite{Eling:2007xh,Yagi:2013ava,Yagi:2013qpa}, dynamical Chern-Simons gravity~\cite{Yunes:2009ch,Yagi:2013mbt,Gupta:2017vsl}, beyond Horndesky theories~\cite{Babichev:2016jom,Sakstein:2016oel}, modify such NS observables, but unfortunately, these modifications are typically degenerate with the nuclear EoS. Yagi and Yunes proposed to solve this problem by finding EoS-insensitive relations between the moment of inertia, the tidal deformability (or Love number) and the quadrupole moment, the so-called ``I-Love-Q'' relations~\cite{Yagi:2013bca,Yagi:ILQ}. Given a measurement of the Love number for a given NS, for example through GW observations, the I-Love-Q relations can be used to infer the moment of inertia or the quadrupole moment. A second independent measurement of either of these two quantities, for example through binary pulsar observations~\cite{Lattimer:2004nj} or observations with the Neutron star Interior Composition ExploreR (NICER)~\cite{Ozel:2015ykl}, then allows an EoS-insensitive test of General Relativity in the strong field regime~\cite{Yagi:2013bca,Yagi:ILQ,Gupta:2017vsl,Doneva:2017jop}.

@@

\begin{figure}
\begin{center} 
\includegraphics[width=\columnwidth]{stackedFisher.eps}
\end{center}
\caption{(Color online) Fisher-estimated statistical uncertainties on the extraction of $\lambda_{0}$ with interferometer (O2, aLIGO, A\texttt{+}, Voyager, CE, ET-D) as a function of the signal-to-noise-ratio expected in each of these instruments, given a single GW170817 detection (circles). The statistical uncertainties with ET are lower than with CE in spite of a lower signal-to-noise ratio because the former is more sensitive above 300 Hz, where tidal effects matter the most (see Sec.~\ref{sec:futureObservations} for further discussion). We also plot the combined statistical uncertainty given $N$ observations consistent with the NS binary merger rate for a 1 year observation (regions), with the top and bottom edges of the regions corresponding to pessimistic and optimistic merger rates. These statistical uncertainties should be compared to the systematic uncertainty on the extraction of $\lambda_{0}$ due to EoS-variability (horizontal dashed line). The statistical and systematic uncertainties cross for Voyager-class detectors. We confirm this conclusion by repeating the statistical analysis with two different waveform models (PhenomD~\cite{PhenomDI,PhenomDII} plus NRTidal corrections~\cite{Dietrich:2017aum,Samajdar:NRTidal} and PhenomD~\cite{PhenomDI,PhenomDII} plus 6PN tidal corrections~\cite{Vines:2011ud,Wade:tidalCorrections}).}
\label{fig:stackedFisher}
\end{figure}

@@

\begin{figure*}
\begin{center} 
\includegraphics[width=.97\columnwidth]{EoSs.eps}
\includegraphics[width=\columnwidth]{hybridEoSs.eps}
\end{center}
\caption{(Color Online) Left: Small representative samples of the unconstrained (dotted) and constrained (solid) sets, together with the 90\% marginalized posterior distribution from the observation of GW170817 (cyan shaded region)~\cite{LIGO:posterior}. There is significantly less variability in the constrained set of EoSs due to the requirement that they be consistent with the GW170817 observation. Right: EoSs for ACS and ACB hybrid stars~\cite{Paschalidis2018}, each transitioning from a hadronic branch (corresponding to a pure hadronic-matter NS) into a quark-matter branch (quark-matter inner core surrounded by hadronic matter) at various transition pressures $P_{\text{tr}}$.
}
\label{fig:eos}
\end{figure*}

@@

\begin{figure*}
\begin{center} 
\includegraphics[width=\columnwidth]{CL.eps}
\includegraphics[width=.96\columnwidth]{RL.eps}
\end{center}
\caption{
(Color online) Similar to Fig.~\ref{fig:ILQ} but for the C-Love (left) and the R-Love relations (right). In the top panels, we show two different fits, one for the unconstrained set and one for the constrained set of EoSs. The bottom panels show the \emph{absolute differences} (rather than fractional difference as in Fig.~\ref{fig:ILQ}) from the fit. The absolute difference is suppressed in the constrained set case relative to the unconstrained set, and results from previous work~\cite{Maselli:2013mva}. In the left panel, we also present the corresponding relations in the hybrid star cases (dashed green) for comparison, where, although still EoS-insensitive, the degree of universality decreases.
}
\label{fig:clove}
\end{figure*}

@@

From the C-Love relations, we can also compute directly the R-Love relation using $R(\Lambda)=M/C(\Lambda)$ for NSs of mass $M$. The right panel of Fig.~\ref{fig:clove} shows the R-Love relations for the constrained and unconstrained sets, with the bottom panel showing the absolute difference of the data and the fits. The C-Love relations allow us to infer the NS radius to better than $\sim 350 \textrm{m} $ in the constrained case, while the error goes up to $1,000 \textrm{m}$ in the unconstrained case, as also tabulated in Table~\ref{tab:maxVar}. The systematic uncertainty in the radius using the constrained fit is thus comparable to the $\sim 140\textrm{m}$ systematic uncertainty in the radius due to the choice of EoS to model the crust~\cite{Gamba:2019kwu}.

@@

\begin{figure}[htb]
\begin{center} 
\includegraphics[width=\columnwidth]{binLoveAbsolute.eps}
\end{center}
\caption{(Color online) Binary Love relations using Eq.~\eqref{eq:binLovefit} fitted to the constrained (dotted maroon) and unconstrained sets (solid green), fixing $q=0.90$, 0.75 and 0.50. The bottom panels show the absolute difference (not the relative difference as in Fig.~\ref{fig:ILQ}) of the fit and the data. The fit to the constrained set shows a reduction in EoS variation relative to both the fit to the unconstrained set and previous work~\cite{Yagi:binLove}. For comparison, we also show the binary Love relations for the 10 hybrid star EoSs (dashed green curves), which present much larger EoS variation. 
}
\label{fig:binLove}
\end{figure}

@@

\begin{figure}
\begin{center} 
\begin{overpic}[width=\columnwidth]{qResiduals.eps}
\put(7,94){\tiny\rotatebox{90}{$\mathcal{R}$}}
\end{overpic}
\begin{overpic}[width=\columnwidth]{LsResiduals.eps}
\put(7,95){\tiny\rotatebox{90}{$\mathcal{R}$}}
\end{overpic}
\end{center}
\caption{Standard deviations of the $\lambda_0$ residuals $\mathcal{R}$ binned in $q$ (top) and $\Lambda_s$ (bottom), highlighting the different error weights across the entire $(q,\Lambda_s)$ parameter space.
In this figure, the violet circles indicate the standard deviation of each bin in $q$-space ($\Lambda_s$-space), while the solid black lines represent the best fit given by Eqs. (\ref{eq:margFit3})-(\ref{eq:margFit4}).
The uncertainty is maximal for $q\sim0.5$ and $\Lambda_s\sim2000$, while it becomes minimal for both low and high values of $q$ and $\Lambda_s$.
Also shown in the figure is the 90th percentile of the un-binned residuals seen in Fig.~\ref{fig:residuals}, taken to be the overall systematic uncertainty introduced by using binary Love relations.
}
\label{fig:qLsResiduals}
\end{figure}

@@

\begin{figure}
\begin{center} 
\includegraphics[width=\columnwidth]{sensitivities.eps}
\end{center}
\caption{(Color online)
Amplitude spectral noise densities $\sqrt{S_n^A(f)}$ plotted for LIGO O2, aLIGO at design sensitivity, A\texttt{+}, Voyager, CE, and ET-D as interpolated from publicly available data.
Spectral noise densities are plotted from $f_{\text{min}}=(23,10,10,7,1,1) \text{ Hz}$, respectively, to $f_{\text{max}}=1649 \text{ Hz}$.
Also shown is $2 \sqrt{f}$ multiplied by the amplitude of the PhenomD~\cite{PhenomDI,PhenomDII} waveform template.
}
\label{fig:sensitivities}
\end{figure}

@@

\begin{figure}
\begin{center} 
\includegraphics[width=\columnwidth]{singleFisherLt.eps}
\end{center}
\caption{(Color Online) Estimated statistical uncertainty $\sigma^A_{\text{GW170817}}$ in the extraction of $\tilde\Lambda$ from a single GW170817-like event as if observed with aLIGO at design sensitivity, A\texttt{+}, Voyager, CE, and ET-D, plotted as a function of the signal-to-noise-ratio $\rho^A_{\text{GW170817}}$ that those detectors would measure for such an event. For comparison, we also plot the systematic uncertainty on $\tilde\Lambda$ due to the use of the binary Love relations. The statistical and the systematic errors become comparable for Voyager-class detectors or better.
}
\label{fig:singleFisherLt}
\end{figure}

>>>>

\author{Zack Carson}
\affiliation{%
 Department of Physics, University of Virginia, Charlottesville, Virginia 22904, USA
}%

\author{Katerina Chatziioannou}
\affiliation{%
 Center for Computational Astrophysics, Flatiron Institute, 162 5th Ave, New York, NY 10010, USA
}%

@@

EoS-insensitive relations can in fact be used for more than just measuring the nuclear EoS. For years, the theoretical physics community considered the possibility of using measurements of NS properties, such as the mass, the radius and the moment of inertia, to constrain deviations from General Relativity in the strong-field regime. Certain modified theories of gravity, such as scalar tensor theories with spontaneous scalarization~\cite{Damour:1996ke}, Einstein-\AE ther and Ho\v rava gravity~\cite{Eling:2007xh,Yagi:2013ava,Yagi:2013qpa}, dynamical Chern-Simons gravity~\cite{Yunes:2009ch,Yagi:2013mbt,Gupta:2017vsl}, beyond Horndesky theories~\cite{Babichev:2016jom,Sakstein:2016oel}, modify such NS observables, but unfortunately, these modifications are typically degenerate with the nuclear EoS~\cite{Pani:2014jra,Minamitsuji:2016hkk,Maselli:2016gxk}. Yagi and Yunes proposed to solve this problem by finding EoS-insensitive relations between the moment of inertia, the tidal deformability (or Love number) and the quadrupole moment, the so-called ``I-Love-Q'' relations~\cite{Yagi:2013bca,Yagi:ILQ}. Given a measurement of the Love number for a given NS, for example through GW observations, the I-Love-Q relations can be used to infer the moment of inertia or the quadrupole moment. A second independent measurement of either of these two quantities, for example through binary pulsar observations~\cite{Lattimer:2004nj} or observations with the Neutron star Interior Composition ExploreR (NICER)~\cite{Ozel:2015ykl}, then allows an EoS-insensitive test of General Relativity in the strong field regime~\cite{Yagi:2013bca,Yagi:ILQ,Gupta:2017vsl,Doneva:2017jop}.

@@

\begin{figure}
\begin{center} 
\includegraphics[width=\columnwidth]{stackedFisher.pdf}
\end{center}
\caption{(Color online) Fisher-estimated statistical uncertainties on the extraction of $\lambda_{0}$ with interferometer (O2, aLIGO, A\texttt{+}, Voyager, CE, ET-D) as a function of the signal-to-noise-ratio expected in each of these instruments, given a single GW170817 detection (circles). The statistical uncertainties with ET are lower than with CE in spite of a lower signal-to-noise ratio because the former is more sensitive above 300 Hz, where tidal effects matter the most (see Sec.~\ref{sec:futureObservations} for further discussion). We also plot the combined statistical uncertainty given $N$ observations consistent with the NS binary merger rate for a 1 year observation (regions), with the top and bottom edges of the regions corresponding to pessimistic and optimistic merger rates. These statistical uncertainties should be compared to the systematic uncertainty on the extraction of $\lambda_{0}$ due to EoS-variability (horizontal dashed line). The statistical and systematic uncertainties cross for Voyager-class detectors. We confirm this conclusion by repeating the statistical analysis with two different waveform models (PhenomD~\cite{PhenomDI,PhenomDII} plus NRTidal corrections~\cite{Dietrich:2017aum,Samajdar:NRTidal} and PhenomD~\cite{PhenomDI,PhenomDII} plus 6PN tidal corrections~\cite{Vines:2011ud,Wade:tidalCorrections}).}
\label{fig:stackedFisher}
\end{figure}

@@

\begin{figure*}
\begin{center} 
\includegraphics[width=.97\columnwidth]{EoSs.pdf}
\includegraphics[width=\columnwidth]{hybridEoSs.pdf}
\end{center}
\caption{(Color Online) Left: Small representative samples of the unconstrained (dotted) and constrained (solid) sets, together with the 90\% marginalized posterior distribution from the observation of GW170817 (cyan shaded region)~\cite{LIGO:posterior}. There is significantly less variability in the constrained set of EoSs due to the requirement that they be consistent with the GW170817 observation. Right: EoSs for ACS and ACB hybrid stars~\cite{Paschalidis2018}, each transitioning from a hadronic branch (corresponding to a pure hadronic-matter NS) into a quark-matter branch (quark-matter inner core surrounded by hadronic matter) at various transition pressures $P_{\text{tr}}$.
}
\label{fig:eos}
\end{figure*}

@@

\begin{figure*}
\begin{center} 
\includegraphics[width=\columnwidth]{CL.pdf}
\includegraphics[width=.96\columnwidth]{RL.pdf}
\end{center}
\caption{
(Color online) Similar to Fig.~\ref{fig:ILQ} but for the C-Love (left) and the R-Love relations (right). In the top panels, we show two different fits, one for the unconstrained set and one for the constrained set of EoSs. The bottom panels show the \emph{absolute differences} (rather than fractional difference as in Fig.~\ref{fig:ILQ}) from the fit. The absolute difference is suppressed in the constrained set case relative to the unconstrained set, and results from previous work~\cite{Maselli:2013mva}. In the left panel, we also present the corresponding relations in the hybrid star cases (dashed green) for comparison, where, although still EoS-insensitive, the degree of universality decreases.
}
\label{fig:clove}
\end{figure*}

@@

From the C-Love relations, we can also compute directly the R-Love relation using $R(\Lambda)=M/C(\Lambda)$ for NSs of mass $M$. The right panel of Fig.~\ref{fig:clove} shows the R-Love relations for the constrained and unconstrained sets, with the bottom panel showing the absolute difference of the data and the fits. The C-Love relations allow us to infer the NS radius to better than $\sim 350 \textrm{m} $ in the constrained case, while the error goes up to $1,000 \textrm{m}$ in the unconstrained case, as also tabulated in Table~\ref{tab:maxVar}. The systematic uncertainty in the radius using the constrained fit is thus comparable to the $\sim 140\textrm{m}$ systematic uncertainty in the radius due to the choice of EoS to model the crust~\cite{Gamba:2019kwu}. 
See also Refs.~\cite{Annala:2017llu,Raithel:2018ncd} for related work on the R-Love relations.

@@

\begin{figure}[htb]
\begin{center} 
\includegraphics[width=\columnwidth]{binLoveAbsolute.pdf}
\end{center}
\caption{(Color online) Binary Love relations using Eq.~\eqref{eq:binLovefit} fitted to the constrained (dotted maroon) and unconstrained sets (solid green), fixing $q=0.90$, 0.75 and 0.50. The bottom panels show the absolute difference (not the relative difference as in Fig.~\ref{fig:ILQ}) of the fit and the data. The fit to the constrained set shows a reduction in EoS variation relative to both the fit to the unconstrained set and previous work~\cite{Yagi:binLove}. For comparison, we also show the binary Love relations for the 10 hybrid star EoSs (dashed green curves), which present much larger EoS variation. 
}
\label{fig:binLove}
\end{figure}

@@

\begin{figure}
\begin{center} 
\begin{overpic}[width=\columnwidth]{qResiduals.pdf}
\put(7,94){\tiny\rotatebox{90}{$\mathcal{R}$}}
\end{overpic}
\begin{overpic}[width=\columnwidth]{LsResiduals.pdf}
\put(7,95){\tiny\rotatebox{90}{$\mathcal{R}$}}
\end{overpic}
\end{center}
\caption{Standard deviations of the $\lambda_0$ residuals $\mathcal{R}$ binned in $q$ (top) and $\Lambda_s$ (bottom), highlighting the different error weights across the entire $(q,\Lambda_s)$ parameter space.
In this figure, the violet circles indicate the standard deviation of each bin in $q$-space ($\Lambda_s$-space), while the solid black lines represent the best fit given by Eqs. (\ref{eq:margFit3})-(\ref{eq:margFit4}).
The uncertainty is maximal for $q\sim0.5$ and $\Lambda_s\sim2000$, while it becomes minimal for both low and high values of $q$ and $\Lambda_s$.
Also shown in the figure is the 90th percentile of the un-binned residuals seen in Fig.~\ref{fig:residuals}, taken to be the overall systematic uncertainty introduced by using binary Love relations.
}
\label{fig:qLsResiduals}
\end{figure}

@@

\begin{figure}
\begin{center} 
\includegraphics[width=\columnwidth]{sensitivities.pdf}
\end{center}
\caption{(Color online)
Amplitude spectral noise densities $\sqrt{S_n^A(f)}$ plotted for LIGO O2, aLIGO at design sensitivity, A\texttt{+}, Voyager, CE, and ET-D as interpolated from publicly available data.
Spectral noise densities are plotted from $f_{\text{min}}=(23,10,10,7,1,1) \text{ Hz}$, respectively, to $f_{\text{max}}=1649 \text{ Hz}$.
Also shown is $2 \sqrt{f}$ multiplied by the amplitude of the PhenomD~\cite{PhenomDI,PhenomDII} waveform template.
}
\label{fig:sensitivities}
\end{figure}

@@

\begin{figure}
\begin{center} 
\includegraphics[width=\columnwidth]{singleFisherLt.pdf}
\end{center}
\caption{(Color Online) Estimated statistical uncertainty $\sigma^A_{\text{GW170817}}$ in the extraction of $\tilde\Lambda$ from a single GW170817-like event as if observed with aLIGO at design sensitivity, A\texttt{+}, Voyager, CE, and ET-D, plotted as a function of the signal-to-noise-ratio $\rho^A_{\text{GW170817}}$ that those detectors would measure for such an event. For comparison, we also plot the systematic uncertainty on $\tilde\Lambda$ due to the use of the binary Love relations. The statistical and the systematic errors become comparable for Voyager-class detectors or better.
}
\label{fig:singleFisherLt}
\end{figure}
